Limma Tutorial

### 2018-04-12 C_R/DESeq2/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

```{r}
library(DESeq2)
library(RColorBrewer)
library(data.table)
library(ggplot2)
library(gplots)

@@

data <- readRDS("../../data/data.Rds")
data <- as.matrix(data)

coldata = data.frame(row.names = colnames(data), virus = unlist(lapply(colnames(data), function (x) substring(unlist(strsplit(x, "[.]"))[1],1,1))), diet = unlist(lapply(colnames(data), function (x) substring(unlist(strsplit(x, "[.]"))[1],2,2))), treatment = unlist(lapply(colnames(data), function (x) unlist(strsplit(x, "[.]"))[1])))

@@

>>>>

```{r}
library(DESeq2)
library(RColorBrewer)
library(data.table)
library(ggplot2)
library(gplots)
library(GGally)

@@

data <- readRDS("../data/data.Rds")
data <- as.matrix(data)

coldata = data.frame(row.names = colnames(data), treatment = unlist(lapply(colnames(data), function (x) unlist(strsplit(x, "[.]"))[1])))

@@

```{r}
ggparcoord(data.frame(log(as.data.frame(assay(rld))/colMeans(as.data.frame(assay(rld))))), columns=1:6, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
```

```{r}
dds <- estimateSizeFactors(dds)
normDat <- counts(dds, normalized=TRUE)
normDat <- as.data.frame(normDat)
ggparcoord(data.frame(log(normDat)/colMeans(normDat)), columns=1:6, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
```

### 2018-04-12 C_R/EdgeR/edgeR-btwnLane/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

---
title: "Honey Bees - edgeR"
author: Lindsay Rutter
output:
  packagedocs::package_docs:
    toc: true
    toc_collapse: true
vignette: >
  %\VignetteEngine{packagedocs::redirect}
---

@@

beeCounts <- readRDS("../../data/data.Rds")

beeCounts <- as.matrix(beeCounts)

@@

Next, I used edgeR normalization.
```{r}
y <- calcNormFactors(y)
```

@@

```{r}
ggparcoord(data.frame(log(y[[1]]/colMeans(y[[1]]))), columns=1:6, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
```

@@

```{r}
allGroups <- c(rep("C",3), rep("T",3))
y$samples$group <- allGroups

@@

```{r}
Group = factor(c(rep("C",3), rep("T",3)))
#targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
design <- model.matrix(~0+Group, data=y$samples)
colnames(design) <- levels(Group)
y <- estimateDisp(y, design)
plotBCV(y)
```

>>>>

---
title: "Honey Bees - edgeR-btwnlane"
author: Lindsay Rutter
output:
  packagedocs::package_docs:
    toc: true
    toc_collapse: true
vignette: >
  %\VignetteEngine{packagedocs::redirect}
---

@@

beeCounts <- readRDS("../../data/data.Rds")
beeCounts <- as.matrix(beeCounts)

@@

Next, I used edgeR normalization.
```{r}
y[[1]] <- betweenLaneNormalization(y[[1]], which="full", round=FALSE)
```

@@

```{r}
ggparcoord(data.frame(log(y[[1]]/colMeans(y[[1]]))), columns=1:24, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
```

@@

```{r}
allGroups <- c(rep("C",12), rep("R",12))
y$samples$group <- allGroups

@@

```{r}
Group = factor(c(rep("C",12), rep("R",12)))
targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
design <- model.matrix(~0+Group, data=y$samples)
colnames(design) <- levels(Group)
y <- estimateDisp(y, design)
plotBCV(y)
```

### 2018-04-12 C_R/EdgeR/edgeR/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

```{r}
Group = factor(c(rep("C",3), rep("T",3)))
#targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
design <- model.matrix(~0+Group, data=y$samples)
colnames(design) <- levels(Group)
y <- estimateDisp(y, design)
plotBCV(y)
```

>>>>

```{r}
Group = factor(c(rep("C",12), rep("R",12)))
design <- model.matrix(~0+Group, data=y$samples)
colnames(design) <- levels(Group)
y <- estimateDisp(y, design)
plotBCV(y)
```

### 2018-04-12 C_R/LimmaVoom/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

group <- factor(c(rep("C",3), rep("T",3)))
x$samples$group <- group
```

Transform and remove low counts.
```{r}
cpm <- cpm(x)
keep.exprs <- rowSums(cpm>1)>=4 # tried filtering up to 24 and not much difference
x <- x[keep.exprs,, keep.lib.sizes=FALSE] # 15,314 to 10,654
dim(x)
x <- calcNormFactors(x, method = "TMM")
```

Make boxplots
```{r}
ggparcoord(data.frame(log(x[[1]]/colMeans(x[[1]]))), columns=1:6, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
```

@@

contr.matrix <- makeContrasts(
   C_T = C-T,
   levels = colnames(design))
```

>>>>

group <- factor(c(rep("C",12), rep("R",12)))
x$samples$group <- group
```

Transform and remove low counts.
```{r}
cpm <- cpm(x)
nRep = length(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])) / length(unique(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])))
keep.exprs <- rowSums(cpm>1)>=nRep # tried filtering up to 24 and not much difference
x <- x[keep.exprs,, keep.lib.sizes=FALSE] # 15,314 to 10,654
dim(x)
x <- calcNormFactors(x, method = "TMM")
```

Make boxplots
```{r}
ggparcoord(data.frame(log(x[[1]]/colMeans(x[[1]]))), columns=1:24, alphaLines=0, boxplot=TRUE, scale="globalminmax") + coord_flip()
```

@@

contr.matrix <- makeContrasts(
   C_R = C-R,
   levels = colnames(design))
```

### 2018-04-12 Manuscript/chapter4.tex
Standard scatterplots Gailbraith data
Redo 2 level factor comparisons D, E, L

Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms. Transcriptomics is one means to achieve this goal. As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees. As far as we know, there are few to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and few to no studies investigating gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense in honeybees.

@@

\subsection{Design of diet experiment}

There are several reasons why, in this part of our previous study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets). First, when assessing diet diversity, a sugar diet is often used as a control. However, such an experimental design does not reflect real-world conditions for honeybees as they rarely face a total lack of pollen (\citealt{DiPasquale}). Second, in studies that compared honeybee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet (\citealt{DiPasquale}). Third, colonies used for pollination in agricultural areas (monoculture) face less diversified pollens (according to Brodschneider, 2010). Pollinating areas are currently undergoing landscape alteration and agriculture intensification, and bees are increasingly faced wtih less diversified diets (monoculture) (\citealt{landscape1}, \citealt{brodschneider}). As a result, there is a need to better understand how monofloral diets affect honeybee health as a step toward mitigating the negative impact of human activity on the honeybee population.

Consequently, in our prior study, for our nutrition factor, we examined two monofloral pollen diets, Cistus (Rockrose) and Castanea (Chestnut). Cistus pollen is generally considered less nutritious than Castanea pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}). For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV. This experimental design resulted in four treatment groups (Cistus pollen without IAPV exposure, Castanea pollen without IAPV exposure, Cistus pollen with IAPV exposure, and Castanea pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.

@@

Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing). A standard Illumina mRNA library was prepared by the DNA facility. Reads were aligned to the BeeBase Version 3.2 genome (\citealt{hbGenome}) from the Hymenoptera Genome Database (\citealt{hymenopteraDB}) using the programs GMAP and GSNAP (\citealt{gsnap}). We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs). We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs). We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honeybees (\citealt{galbraith}). We tested each DEG analysis using recommended parameters with DESeq2 (\citealt{deseq2}), edgeR (\citealt{edger}), and LimmaVoom (\citealt{limma}). In all cases, we used a false discovery rate (FDR) threshold of 0.05 (\citealt{benjamini}). Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines).

>>>>

Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality. For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on energy availability and vigor (resilience). Transcriptomics is one means to achieve this goal. Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection. This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a "buffer" against other stressors (\citealt{AdamTothReview}). As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees. As far as we know, there are few to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honeybee gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense.

@@

\subsection{Design of two-factor experiment}

There are several reasons why we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets). First, when assessing diet diversity, a sugar diet is often used as a control. However, such an experimental design does not reflect real-world conditions for honeybees as they rarely face a total lack of pollen (\citealt{DiPasquale}). Second, in studies that compared honeybee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet (\citealt{DiPasquale}). Third, colonies used for pollination in agricultural areas (monoculture) face less diversified pollens (according to Brodschneider, 2010). Pollinating areas are currently undergoing landscape alteration and agriculture intensification, and bees are increasingly faced with less diversified diets (monoculture) (\citealt{landscape1}, \citealt{brodschneider}). As a result, there is a need to better understand how monofloral diets affect honeybee health as a step toward mitigating the negative impact of human activity on the honeybee population.

Consequently, for our nutrition factor, we examined two monofloral pollen diets, Cistus (Rockrose) and Castanea (Chestnut). Cistus pollen is generally considered less nutritious than Castanea pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}). For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV. This experimental design resulted in four treatment groups (Cistus pollen without IAPV exposure, Castanea pollen without IAPV exposure, Cistus pollen with IAPV exposure, and Castanea pollen with IAPV exposure) that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.

@@

Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing). A standard Illumina mRNA library was prepared by the DNA facility. Reads were aligned to the BeeBase Version 3.2 genome (\citealt{hbGenome}) from the Hymenoptera Genome Database (\citealt{hymenopteraDB}) using the programs GMAP and GSNAP (\citealt{gsnap}). We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs). We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs). We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honeybees (\citealt{galbraith}). We tested each DEG analysis using recommended parameters with DESeq2 (\citealt{deseq2}), edgeR (\citealt{edger}), and LimmaVoom (\citealt{limma}). In all cases, we used a false discovery rate (FDR) threshold of 0.05 (\citealt{benjamini}). Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines). The \pkg{eulerr} shiny application was used to construct Venn diagram overlap images (\citealt{euler}).

### 2018-04-12 N_V/EdgeR/edgeR/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

```{r}
Group = factor(c(rep("N",12), rep("V",12)))
targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
design <- model.matrix(~0+Group, data=y$samples)
colnames(design) <- levels(Group)
y <- estimateDisp(y, design)
plotBCV(y)
```

>>>>

```{r}
Group = factor(c(rep("N",12), rep("V",12)))
design <- model.matrix(~0+Group, data=y$samples)
colnames(design) <- levels(Group)
y <- estimateDisp(y, design)
plotBCV(y)
```

### 2018-04-12 N_V/LimmaVoom/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

group <- factor(c(rep("N",12), rep("V",12)))
x$samples$group <- group
lane <- as.factor(c("L12","L12","L12","L12","L34","L34","L12","L12","L34","L34","L34","L34","L12","L12","L12","L12","L34","L34","L12","L12","L34","L34","L34","L34"))
x$samples$lane <- lane
```

Transform and remove low counts.
```{r}
cpm <- cpm(x)
keep.exprs <- rowSums(cpm>1)>=4 # tried filtering up to 24 and not much difference
x <- x[keep.exprs,, keep.lib.sizes=FALSE] # 15,314 to 10,654
dim(x)
x <- calcNormFactors(x, method = "TMM")
```

@@

Create MDS plots
```{r}
library(RColorBrewer)
lcpm <- cpm(x, log=TRUE)
par(mfrow=c(1,2))
col.group <- group
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)
col.lane <- lane
levels(col.lane) <-  brewer.pal(nlevels(col.lane), "Set2")
col.lane <- as.character(col.lane)
```

@@

```{r}
plotMDS(lcpm, labels=colnames(lcpm), col=col.lane, dim=c(3,4))
```

```{r}
glMDSPlot(lcpm, labels=paste(group, lane, colnames(lcpm), sep="_"), groups=x$samples[,c(1,4)], launch=FALSE)
```

Create design matrix. There are many ways to setup a design matrix. Here, we removed the intercept from group (the first factor), but kept the intercept from lane. This allows us to do contrasts with group more easily.
```{r}
design <- model.matrix(~0+group+lane)
colnames(design) <- gsub("group", "", colnames(design))

>>>>

group <- factor(c(rep("N",12), rep("V",12)))
x$samples$group <- group
```

Transform and remove low counts.
```{r}
cpm <- cpm(x)
nRep = length(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])) / length(unique(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])))
keep.exprs <- rowSums(cpm>1)>=nRep # tried filtering up to 24 and not much difference
x <- x[keep.exprs,, keep.lib.sizes=FALSE] # 15,314 to 10,654
dim(x)
x <- calcNormFactors(x, method = "TMM")
```

@@

Create MDS plots
```{r}
library(RColorBrewer)
lcpm <- cpm(x, log=TRUE)
par(mfrow=c(1,2))
col.group <- group
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)
```

@@

Create design matrix. There are many ways to setup a design matrix. Here, we removed the intercept from group (the first factor). This allows us to do contrasts with group more easily.
```{r}
design <- model.matrix(~0+group)
colnames(design) <- gsub("group", "", colnames(design))

### 2018-04-12 VirusHoneyBee/EdgeR/edgeR/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

```{r}
Group = factor(c(rep("C",3), rep("T",3)))
#targets <- data.frame(rownames=paste0("Sample",1:24), Group = Group)
design <- model.matrix(~0+Group, data=y$samples)
colnames(design) <- levels(Group)
y <- estimateDisp(y, design)
plotBCV(y)
```

>>>>

```{r}
Group = factor(c(rep("C",3), rep("T",3)))
design <- model.matrix(~0+Group, data=y$samples)
colnames(design) <- levels(Group)
y <- estimateDisp(y, design)
plotBCV(y)
```

### 2018-04-12 VirusHoneyBee/LimmaVoom/makeDataMetrics.Rmd
Redo 2 level factor comparisons D, E, L

Transform and remove low counts.
```{r}
cpm <- cpm(x)
keep.exprs <- rowSums(cpm>1)>=4 # tried filtering up to 24 and not much difference
x <- x[keep.exprs,, keep.lib.sizes=FALSE] # 15,314 to 10,654
dim(x)
x <- calcNormFactors(x, method = "TMM")
```

>>>>

Transform and remove low counts.
```{r}
cpm <- cpm(x)
nRep = length(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])) / length(unique(lapply(colnames(beeCounts), function (x) unlist(strsplit(x, "[.]"))[1])))
keep.exprs <- rowSums(cpm>1)>=nRep # tried filtering up to 24 and not much difference
x <- x[keep.exprs,, keep.lib.sizes=FALSE] # 15,314 to 10,654
dim(x)
x <- calcNormFactors(x, method = "TMM")
```
